Make any stupid plot you want

### 2021-01-20 draft/sections/math.tex
rotation group sketch

\subsubsection{Variables: Fiber Space $F$}
\label{sec:data_fiber}
The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset. For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $\{tp_0, \ldots, tp_n\} \subset \mathbb{R}$. Lets rename this $F$ to a component 
\begin{equation}
F_0 = \{tp_0, \ldots, tp_n\} 
\end{equation}
so that more variables can be added to the fiber. Then lets add the time from GHCN $F_1 =\{t_0, \ldots, t_n\}$
\begin{equation}
F =  \{tp_0, \ldots, tp_n\}  \times \{ti_0, \ldots, ti_n\} 
\end{equation}

The elements in a given fiber set do not need to be scalers; for example the location can be encoded as the set of tuples $(\textit{lat},\textit{lon})$.  We add the location to the fiber 
\begin{equation}
    F = \{tp_0, \ldots, tp_n\}  \times \{ti_0, \ldots, ti_n\} \times 
    \{(la_0, ln_0),\ldots, (la_j, ln_k), \ldots (la_n,ln_n)\}
\end{equation}
such that one record $r$ in the fiber would have the form $(tp, t, (la,ln))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
\begin{align}
    F_0 &= (\textit{temperature}, \{tp_0, \ldots, tp_n\}, M_0)\\
    F_1 &= (\textit{time}, \{ti_0, \ldots, ti_n\}, M_1)\\
    F_2 &= (\textit{location},   \{(la_0, ln_0),\ldots, (la_j, ln_k), \ldots (la_n,ln_n)\}, M_2)
\end{align}
 or more generally (name, set, monoid)
 \begin{equation}
    F_i = (\textit{name}, \{x_{i}, \ldots, x_{n}\}, M_{i})
 \end{equation}
as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
\begin{equation}
F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
\end{equation}
and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:data_monoid}
A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or rotating a shape; this model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}. One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.

We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$. A monoid \cite{Monoid2021} $M$ is a set with an associative binary operator $\ast:M \times M\rightarrow M$. A monoid has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$. A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in F, e\in M,  e\bullet x = x 
\end{align*}

For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $F_0$ are the set of all transformations $tp_i + c$ such that 
\begin{equation}
F_0 = (\textit{temperature}, \{tp_0, \ldots, tp_n\}, \{tp_0 + c_0, \ldots, tp_0 + c_n, \ldots, tp_n + c_0, \ldots, tp_n + c_n\} ) 
\end{equation}

@@

\begin{figure}[ht!]
    \includegraphics[width=\textwidth]{figures/math/k_different_types.png}
    \label{fig:base_space_types}
    \caption{The topological base space $K$ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
\end{figure}

@@

We encode indexing from $K$ to $E$ as sections $\tau$ of the fiber bundle that return corresponding records. For example when $K$ has
\begin{align*}
    \textbf{vertexes}\;& \tau(k) \text{k= vertex id}
    \textbf{edges}\; \tau(k,\alpha), \text{k = edge id, $\alpha$ = distance along edge} 
    \textbf{faces}\; \tau(k,\alpha_1, \alpha_2), \text{k = face id, $\alpha_1$ = x on face, $\alpha_2$ = y on face}
\end{align*}
such that we can check for continuity of sections $\tau$ by verifying that the edges align such that $\tau(k_{i},1) = \tau(k_{j},0)$.

@@

which is a decomposition of $F$. The $K$ remains the same because the connectivity of records does not change just because there are fewer variables in each record.

\subsubsection{Values: Sections $\tau$}
\label{sec:data_section}
%%% visual variable encoding in a consistent/atomic way in a complex structure 
\label{sec:section_data}
The sections of the fiber bundles are the instances of the data. They are the functions that map from a point $k \in K$ to a set of values in $F$. For example, if we have a database, the section is the table that yields a row given an index. The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$ 
\begin{equation}
    \label{eq:section}
    \begin{tikzcd}
        F \arrow[r, hook] & E \arrow[d, "\pi"']           \\
        & K \arrow[u, "\tau"', bend right]
    \end{tikzcd}
\end{equation}

%%% change to b\c locally trivial, can represent 
such that in a trivial fiber bundle, $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
\begin{equation}
    \label{eq:section_return}
    \tau(k) = (k, (x_{F_{0}}, \ldots, x_{F_{n}}))
\end{equation}

%%%can decompose bundles into Ei such that \tau: E_{i} \rightarrow K, where F_{i} is the fiber of E_{i}
which we can also decompose such that
\begin{align}
\tau_{0}(k) &= (k, (x_{F_{0}}))\\
            & \vdotswithin{=}\\
\tau_{n}(k) &= (k, (x_{F_{n}}))
\end{align}

@@

%% the stalk contains all the local information, typically we only need some of the information, such as the firtst or second direivative. in practice we only need the jet which is...in practive will extend the fiber 
The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is a for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.

@@

\subsubsection{Base space}
The underlying topology $S$ of a graphic often needs more dimensions than the data topology $K$ because of the specifications of the display space. For example, a line plot on a plane (such as a screen or a piece of paper) by necessity needs to also have a thickness so that it is visible, which maps back to a set of connected points in $H$. The topology of these connected points is therefore the region $s \subset S$ such that $\xi: S \rightarrow K$ is a deformation retraction \cite{RetractionTopology2020}
\begin{equation}
    \begin{tikzcd}
        E \arrow[d, "\pi"'] & H \arrow[d, "\pi"'] \\
        K                   & S \arrow[l, "\xi"']
        \end{tikzcd}
\end{equation}

@@

\paragraph{Example}
For a physical screen display, we can consider a predisplay space that is a trivial fiber bundle $H=\mathbb{R}^{5}\times S$ such that $\rho$ is
\begin{equation}
    \rho(s)  = [x(s), y(s), r(s), g(s), b(s)]
    \label{eq:rho}
\end{equation}

To draw an image, a region, $H$ is inverse mapped into a region $s \in S$ where
\begin{equation}
s = \rho^{-1}_{\tiny{XY}}(xy)
\end{equation}
such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields:
\begin{align}
    r &= \iint\limits_s \rho_{\tiny{R}}(s)ds^{2}\\
    g &= \iint\limits_s \rho_{\tiny{G}}(s)ds^{2}\\
    b &= \iint\limits_s \rho_{\tiny{B}}(s)ds^{2}
\end{align}

@@

In this section we will define the artist as a mapping from a sheaf $\mathcal{O}(E)$  to $\mathcal{O}(H)$. 
\begin{equation}
    A: \mathcal{O}(E) \rightarrow \mathcal{O}(H)
\end{equation}

@@

\begin{equation}
    \label{eq:artist_daigram}
    \begin{tikzcd}
        E \arrow[r, "\nu"] \arrow[rd, "\pi"'] & V \arrow[d, "\pi"] & \xi^*V \arrow[r, "Q"] \arrow[d, "\xi^*\pi"'] \arrow[l, "\xi^*"'] & H \arrow[ld, "\pi"] \\
                                              & K                  & S \arrow[l, "\xi"']                                              &                    
        \end{tikzcd}
\end{equation}
The pullback map $\xi^*$ copies each value in $V$ over $k$ to $s$ in corresponding $S_k$ such that $\xi^*V$ can have multiple values that map to one value in $V$.

@@

\subsubsection{Assembling Marks}
\label{sec:artist_q}
\begin{figure}[ht!]
    \includegraphics[width=\textwidth]{figures/math/path_of_q}
    \label{fig:artist_q}
    \caption{The $\nu$ functions convert data $E$ to visual $V$. $Q$ assembles the different types of visual parameters $V_{i}$ into a graphic in $H$. $Q\circ\mu(\xi^{-1}J)$ forms a visual mark by applying $Q$ to a region mapped to connected components $J \subset K$.}  
\end{figure}

>>>>

\subsubsection{Variables: Fiber Space $F$}
\label{sec:data_fiber}
The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset. For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ \mathscr{}\subset \mathbb{R}$. Lets rename this $F$ to a component 
%%use spivaks U w/ subset data {x \in R | x >= 0}
\begin{equation}
F_0 = \{T_0, \ldots, T_n\} 
\end{equation}
so that more variables can be added to the fiber. Then lets add the time from GHCN $F_1 =\{t_0, \ldots, t_n\}$
\begin{equation}
F =  \{T_0, \ldots, T_n\}  \times \{t_0, \ldots, t_n\} 
\end{equation}

The elements in a given fiber set do not need to be scalers; for example the location can be encoded as the set of tuples $(\textit{lat},\textit{lon})$.  We add the location to the fiber 
%lat,lon in U
\begin{equation}
    F = \{T_0, \ldots, T_n\}  \times \{t_0, \ldots, t_n\} \times 
    \{(a_0, o_0),\ldots, (a_j, o_k), \ldots (a_n, o_n)\}
\end{equation}
such that one record $r$ in the fiber would have the form $(T, t, (a, o))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
\begin{align}
    F_0 &= (\textit{temperature}, \{tp_0, \ldots, tp_n\}, M_0)\\
    F_1 &= (\textit{time}, \{ti_0, \ldots, ti_n\}, M_1)\\
    F_2 &= (\textit{location},   \{(la_0, ln_0),\ldots, (la_j, ln_k), \ldots (la_n,ln_n)\}, M_2)
\end{align}
 or more generally (name, set, monoid)
 \begin{equation}
    F_i = (\textit{name}, \{x_{i}, \ldots, x_{n}\}, M_{i})
 \end{equation}
as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
\begin{equation}
F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
\end{equation}
and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:data_monoid}
A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or enlarging a shape. This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.

\begin{figure}
    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
    \label{fig:data_monoid_rotation}
\end{figure}
One such set of transforms is the rotation group showin in figure~\ref{fig:}
One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.

% change to Mi 
We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$. A monoid \cite{Monoid2021} $M$ is a set with an associative binary operator $\ast:M \times M\rightarrow M$. A monoid has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$. A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
\begin{align*}%lost the \ast
    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in F, e\in M,  e\bullet x = x 
\end{align*}

For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $F_0$ are the set of all transformations $tp_i + c$ such that %%generalize out to x in \U_{temp}, c in \R
%%the action of the interval is the + 
%% c is the element that shifts from x1 to x2
\begin{equation}
F_0 = (\textit{temperature}, \{tp_0, \ldots, tp_n\}, \{tp_0 + c_0, \ldots, tp_0 + c_n, \ldots, tp_n + c_0, \ldots, tp_n + c_n\} ) 
\end{equation}

@@

\begin{figure}[ht!]
    \includegraphics[width=.5\textwidth]{figures/math/k_different_types.png}
    \label{fig:base_space_types}
    \caption{The topological base space $K$ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
\end{figure}

@@

We encode indexing from $K$ to $E$ as sections $\tau$ of the fiber bundle that return corresponding records.

%%if we had a coordinate on

%%pull this down to triangulization section 
For example when $K$ has
\begin{align*}
    \textbf{vertexes}\;& \tau(k) \text{k= vertex id}\\
    \textbf{edges}\;& \tau(k,\alpha), \text{k = edge id, $\alpha$ = distance along edge} \\
    \textbf{faces}\;& \tau(k,\alpha_1, \alpha_2), \text{k = face id, $\alpha_1$ = x on face, $\alpha_2$ = y on face}
\end{align*}
such that we can check for continuity of sections $\tau$ by verifying that the edges align such that $\tau(k_{i},1) = \tau(k_{j},0)$.

\begin{figure}
    %%include transition function sketch
\end{figure}

We can represent more complex spaces with transition functions that are maps from the fiber to the fiber along the edges. can also fold in if monoid is commutative on transition function:

@@

which is a decomposition of $F$. The $K$ remains the same because the connectivity of records does not change just because there are fewer elements in each record.

\subsubsection{Values: Sections $\tau$}
\label{sec:data_section}
%%maybe pull down/redo figure
%%% visual variable encoding in a consistent/atomic way in a complex structure 
\label{sec:section_data}
The sections of the fiber bundles are the instances of the data. They are the functions that map from a point $k \in K$ to a set of values in $F$. For example, if we have a database, the section is the table that yields a row given an index. The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$ 
\begin{equation}
    \label{eq:section}
    \begin{tikzcd}
        F \arrow[r, hook] & E \arrow[d, "\pi"']           \\
        & K \arrow[u, "\tau"', bend right]
    \end{tikzcd}
\end{equation}
such that %%insrt math equiality here 
%%% change to b\c locally trivial, can represent 
If the fiber bundle is trivial, then $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
\begin{equation}
    \label{eq:section_return}
    \tau(k) = (k, (x_{F_{0}}, \ldots, x_{F_{n}}))
\end{equation}
which is also true when the fiber bundle is locally trivial over $U \subset K$.
%%%can decompose bundles into Ei such that \tau: E_{i} \rightarrow K, where F_{i} is the fiber of E_{i}
which we can also decompose such that
\begin{align}
\tau_{0}(k) &= (k, (x_{F_{0}}))\\
            & \vdotswithin{=}\\
\tau_{n}(k) &= (k, (x_{F_{n}}))
\end{align}

@@

%% the stalk contains all the local information, typically we only need some of the information, such as the firtst or second direivative. in practice we only need the jet which is...in practive will extend the fiber /fatten the E , notatoe E' = E + Jet
The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.

@@

%%define a region b = [top, bottom, right, left]
%% inverse of b is capital S_b
\subsubsection{Base space}
The underlying topology $S$ of a graphic often needs more dimensions than the data topology $K$ because of the specifications of the display space. For example, a line plot on a plane (such as a screen or a piece of paper) by necessity needs to also have a thickness so that it is visible, which maps back to a set of connected points in $H$. The topology of these connected points is therefore the region $s \subset S$ such that $\xi: S \rightarrow K$ is a deformation retraction \cite{RetractionTopology2020}
\begin{equation}
    \begin{tikzcd}
        E \arrow[d, "\pi"'] & H \arrow[d, "\pi"'] \\
        K                   & S \arrow[l, "\xi"']
        \end{tikzcd}
\end{equation}

@@

\paragraph{Example}
For a physical screen display, we can consider a predisplay space that is a trivial fiber bundle $H=\mathbb{R}^{5}\times S$ such that $\rho$ is
\begin{equation}
    \rho(s)  = (D_x(s), D_y(s), D_r(s), D_g(s), D_b(s))
    \label{eq:rho}
\end{equation}

To draw an image, a region, $H$ is inverse mapped into a region $s \in S$ where %%% 
\begin{equation}
s = \rho^{-1}_{\tiny{XY}}(xy)
\end{equation}
such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields
%% redefining notation 
\begin{align}
    R_b &= \iint\limits_s D_r(s)ds^{2}\\
    G_b &= \iint\limits_s D_g(s)ds^{2}\\
    B_b &= \iint\limits_s D_b(s)ds^{2}
\end{align}

@@

%% E' not E
In this section we will define the artist as a mapping from a sheaf $\mathcal{O}(E)$  to $\mathcal{O}(H)$. 
\begin{equation}
    A: \mathcal{O}(E) \rightarrow \mathcal{O}(H)
\end{equation}

@@

\begin{equation}
    \label{eq:artist_diagram}
    \begin{tikzcd}
        E \arrow[r, "\nu"] \arrow[rd, "\pi"'] & V \arrow[d, "\pi"] & \xi^*V \arrow[r, "Q"] \arrow[d, "\xi^*\pi"'] \arrow[l, "\xi^*"'] & H \arrow[ld, "\pi"] \\
                                              & K                  & S \arrow[l, "\xi"']                                              &                    
        \end{tikzcd}
\end{equation}
The pullback map $\xi^*$ copies each value in $V$ over $k$ to $s$ in corresponding $S_k$ such that $\xi^*V$ can have multiple values that map to one value in $V$.

@@

\subsubsection{Assembling Marks}
\label{sec:artist_q}
\begin{figure}[ht!]
    \includegraphics[width=\textwidth]{figures/math/path_of_q}
    \label{fig:artist_q}
    \caption{The $\nu$ functions convert data $E$ to visual $V$. $Q$ assembles the different types of visual parameters $V_{i}$ into a graphic in $H$. $Q\circ\mu(\xi^{-1}J)$ forms a visual mark by applying $Q$ to a region mapped to connected components $J \subset K$.}
    %%maybe change to \tau, \mu, \rho  
\end{figure}
